Practical Machine Learning course work
========================================================

The definition of the problem is given here: ***https://class.coursera.org/predmachlearn-002/human_grading/view/courses/972090/assessments/4/submissions***

First we import the needed libraries:
```{r}
library(randomForest)
library(caret)
```
Then we read the data to corresponding data frames: 

```{r}
setwd("D:/Files/Dropbox/Studies/Data Science/04 Machine Learning/coursework")
training <- read.csv("pml-training.csv")
testing <- read.csv("pml-testing.csv")

```
Now we analyze what's inside the data:

```{r}
names(training) # 160 columns, last column: classe
nrow(training) # 19622 rows
names(testing) # 160 columns, last column: problem_id
nrow(testing) # 20 rows
table(training$classe) # all classes have values, fairly evenly distributed
sum(is.na(training$classe)) # no na values
nrow(training[!complete.cases(training),]) # 19216 rows - vast majority have na values!
sum(is.na(training)) # 1287472
##trainingCC <- training[complete.cases(training),]
##trainingCC$classe<-as.factor(trainingCC$classe)
```

Now let's get rid of columns that are mostly NAs
```{r}
drop_names<-c('kurtosis_roll_belt',
'kurtosis_picth_belt',
'kurtosis_yaw_belt',
'skewness_roll_belt',
'skewness_roll_belt.1',
'skewness_yaw_belt',
'max_roll_belt',
'max_picth_belt',
'max_yaw_belt',
'min_roll_belt',
'min_pitch_belt',
'min_yaw_belt',
'amplitude_roll_belt',
'amplitude_pitch_belt',
'amplitude_yaw_belt',
'var_total_accel_belt',
'avg_roll_belt',
'stddev_roll_belt',
'var_roll_belt',
'avg_pitch_belt',
'stddev_pitch_belt',
'var_pitch_belt',
'avg_yaw_belt',
'stddev_yaw_belt',
'var_yaw_belt',
'var_accel_arm',
'avg_roll_arm',
'stddev_roll_arm',
'var_roll_arm',
'avg_pitch_arm',
'stddev_pitch_arm',
'var_pitch_arm',
'avg_yaw_arm',
'stddev_yaw_arm',
'var_yaw_arm',
'kurtosis_roll_arm',
'kurtosis_picth_arm',
'kurtosis_yaw_arm',
'skewness_roll_arm',
'skewness_pitch_arm',
'skewness_yaw_arm',
'max_roll_arm',
'max_picth_arm',
'max_yaw_arm',
'min_roll_arm',
'min_pitch_arm',
'min_yaw_arm',
'amplitude_roll_arm',
'amplitude_pitch_arm',
'amplitude_yaw_arm',
'kurtosis_roll_dumbbell',
'kurtosis_picth_dumbbell',
'kurtosis_yaw_dumbbell',
'skewness_roll_dumbbell',
'skewness_pitch_dumbbell',
'skewness_yaw_dumbbell',
'max_roll_dumbbell',
'max_picth_dumbbell',
'max_yaw_dumbbell',
'min_roll_dumbbell',
'min_pitch_dumbbell',
'min_yaw_dumbbell',
'amplitude_roll_dumbbell',
'amplitude_pitch_dumbbell',
'amplitude_yaw_dumbbell',
'var_accel_dumbbell',
'avg_roll_dumbbell',
'stddev_roll_dumbbell',
'var_roll_dumbbell',
'avg_pitch_dumbbell',
'stddev_pitch_dumbbell',
'var_pitch_dumbbell',
'avg_yaw_dumbbell',
'stddev_yaw_dumbbell',
'var_yaw_dumbbell',
'kurtosis_roll_forearm',
'kurtosis_picth_forearm',
'kurtosis_yaw_forearm',
'skewness_roll_forearm',
'skewness_pitch_forearm',
'skewness_yaw_forearm',
'max_roll_forearm',
'max_picth_forearm',
'max_yaw_forearm',
'min_roll_forearm',
'min_pitch_forearm',
'min_yaw_forearm',
'amplitude_roll_forearm',
'amplitude_pitch_forearm',
'amplitude_yaw_forearm',
'var_accel_forearm',
'avg_roll_forearm',
'stddev_roll_forearm',
'var_roll_forearm',
'avg_pitch_forearm',
'stddev_pitch_forearm',
'var_pitch_forearm',
'avg_yaw_forearm',
'stddev_yaw_forearm',
'var_yaw_forearm')

trainingsml<-training[,!(names(training) %in% drop_names)]

subtrn <- createDataPartition(y=trainingsml$classe, p=0.05, list=F) # We're building RF on a small subset of training rows because they are enough

nrow(subtrn)

trsm1 <- trainingsml[subtrn, ]
table(factor(training$classe))

```

Next we will fit a random forrest model

```{r}
#rfmodel<- randomForest(classe~., data=trainingCC[,-c(1,2)], prox=TRUE, importance=T)
rfModel<- train(classe~., data=trsm1[,-c(1,2)], method="rf", prox=TRUE,)
rfModel
rfModel2<- train(classe~., data=trainingsml[,-c(1,2)], method="rf", prox=TRUE, trControl = trainControl(method = "cv", number = 2))
```

Now we will predict values

```{r}
pred <- predict(rfModel, testing)
testing$classe <- pred

```

Now we will write files for the submission

```{r}

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    print(c("Writing file #", i))
  }
}

pml_write_files(pred)
```




